Polynomial phase estimation by phase unwrapping

### 2013-02-21 paper.tex
added minor

\begin{proposition}\label{prop:lattice-theory-constructing_a_rectangular_tesselating_region}
Let  $\Lambda$ be an $n$ dimensional lattice and $\Bbf \in \reals^{n\times n}$ be a generator matrix for $\Lambda$. Let $\Bbf = \Qbf\Rbf$ where $\Qbf$ is orthonormal and $\Rbf$ is upper triangular with elements $r_{ij}$.  Then the rectangular prism $\Qbf P$ where
\[
P = \prod_{k=1}^{n}{\left[-\frac{r_{kk}}{2}, \frac{r_{kk}}{2}\right)}
\]
is a set of coset representatives for $\reals^n / \Lambda$.
\end{proposition}
\begin{IEEEproof}
This result is well known~\cite[Chapter IX, Theorem IV]{Cassels_geom_numbers_1997}~\cite[Proposition 2.1]{McKilliam2010thesis}.  This result is for lattices with full rank.  A result in the general case can be obtained similarly, but is not required here.  
\end{IEEEproof}

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}. The nearest lattice point problem is: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find the lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, solve
\[
\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.
\]
The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself, or rather a basis thereof, is considered as an additional input parameter \cite{micciancio_hardness_2001, AjtaiShortestVecProbNPHard1998}.  Nevertheless, algorithms exist that can compute the nearest lattice point in reasonable time if the dimension is small \cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. One such algorithm introduced by Pohst \cite{Pohst_sphere_decoder_1981} in 1981 was popularised in signal processing and communications fields by Viterbo and Boutros \cite{Viterbo_sphere_decoder_1999} and has since been called the \emph{sphere decoder}.  With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.  Approximate algorithms for computing the nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986} which has worst case polynomial complexity.  Yet another approximate algorithm is the $K$-best algorithm studied in~\cite{Zhan2006_K_best_sphere_decoder} that operates similarly to the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} used in coding theory.  The $K$-best algorithm is more expensive that Babai's algorithm, but typically produces a better approximation of the nearest lattice point.  In this paper we make use of the $K$-best algorithm for approximating the LSU estimator.  We will find that is produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount.   We will not detail the workings of this algorithm as an excellent description has been given in~\cite{Zhan2006_K_best_sphere_decoder}.

@@

Given a polynomial $g(x) = a_0 + a_1x + \dots + a_m x^m$, let
\[
\coef(g) = \left[ \begin{array}{ccccc} a_0 & a_1 & a_2 & \dots & a_m \end{array} \right]^\prime,
\]
where superscript $^\prime$ indicates the transpose, denote the column vector of length $m+1$ containing the coefficients of $g$.  If $y$ and $z$ differ by a polynomial from $\mathcal{Z}$ then we can write $y = z + p$ where $p \in \mathcal{Z}$ and then also $\coef(y) = \coef(z) + \coef(p)$.
%\footnote{In group theory terminology $\coef(\cdot)$ coupled with vector addition is called a \term{group homomorphism}.\index{group homomorphism}}. 
  Consider the set
\[
L_{m+1} = \{ \coef(p) \mid p \in \mathcal{Z} \},
\]
containing the coefficient vectors corresponding to the polynomials in $\mathcal{Z}$.  Since the integer valued polynomials are a basis for $\mathcal{Z}$,
\begin{align*}
L_{m+1} &= \{ \coef(c_0 p_0 + c_1p_1 + \dots + c_mp_m) \mid c_i \in \ints \} \\
&= \{ c_0 \coef(p_0) + \dots + c_m\coef(p_m) \mid c_i \in \ints \}.
\end{align*}
Let
\[
\Pbf = \left[ \begin{array}{cccc} \coef(p_0)& \coef(p_1)& \dots& \coef(p_m)  \end{array} \right]
\]
be the $m+1$ by $m+1$ matrix with columns given by the coefficients of the integer valued polynomials.  Then,
\[
L_{m+1} = \{ \xbf = \Pbf\ubf \mid \ubf \in \ints^{m+1} \}
\]
and it is clear that $L_{m+1}$ is an $m+1$ dimensional lattice.  That is, the set of coefficients of the polynomials from $\mathcal{Z}$ forms a lattice with generator matrix $\Pbf$. We can restate Theorem~\ref{thm:circpolysampledthm} as:
\begin{corollary}\label{cor:circpolysampledcoef}
Two polynomial phase signals $s(t) = e^{2\pi j y(t)}$  and $r(t) = e^{2\pi j z(t)}$ are aliases if and only if $\coef(y)$ and $\coef(z)$ differ by a lattice point in $L_{m+1}$.
\end{corollary}

@@

As an example consider the polynomial phase signal of order zero $e^{2\pi j \mu_0}$.  Since $e^{2\pi j \mu_0} = e^{2\pi j(\mu_0 + k)}$ for any integer $k$ we must, in order to ensure identifiability, restrict $\mu_0$ to some interval of length $1$.  A natural choice is the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$. The lattice $L_1$ is the 1-dimensional integer lattice $\ints$ and the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$ corresponds to the Voronoi cell of $L_1$. 
When $m=1$ it turns out that a natural choice of identifiable region is the square box $[-\nicefrac{1}{2}, \nicefrac{1}{2})^2$. This corresponds with the \term{Nyquist criterion}.  The lattice $L_2$ is equal to $\ints^2$ so the box $[-\nicefrac{1}{2}, \nicefrac{1}{2})^2$ corresponds with the Voronoi cell of $L_2$.  
When $m > 1$ the identifiable region becomes more complicated and $L_{m+1} \neq \ints^{m+1}$. %However, we can always construct an identifiable region by taking a tessellating region of lattice $L_{m+1}$.  %Two convenient tessellating regions are the Voronoi cell of $L$ and the rectangular tessellating region given by~\eqref{eq:rectangular_identifiable_region}.

In general there are infinitely many choices for the identifiable region. A natural choice is the Voronoi cell of $L_{m+1}$ used in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}. Another potential choice is a fundamental parallelepiped of $L_{m+1}$. In this paper we will use the rectangular set constructed using Proposition~\ref{prop:lattice-theory-constructing_a_rectangular_tesselating_region}. Observe that $\Pbf$ is upper triangular with $k$th diagonal element equal to $\tfrac{1}{k!}$.  So this rectangular region is
\begin{equation}\label{eq:rectangular_identifiable_region}
B = \prod_{k=0}^{m}\left[ -\frac{0.5}{k!}, \frac{0.5}{k!}  \right).
\end{equation}
We will make use of this region when deriving the statistical properties of the LSU estimator in the next section.

Given vectors $\xbf$ and $\ybf$ in $\reals^{m+1}$ we say that $\xbf \equiv \ybf \bmod L_{m+1}$ if $\xbf$ and $\ybf$ differ by a lattice point in $L_{m+1}$.  We define the function $\dealias(\xbf)$ to take $\xbf$ to its coset representative inside $B$. That is, $\dealias(\xbf) = \zbf \in B$ where $\xbf - \zbf \in L_{m+1}$.  %The effect of $\dealias(\xbf)$ is to \emph{dealias} the polynomial phase coefficients. 
When $m = 0$ or $1$ then $\dealias(\xbf) = \fracpart{\xbf}$ where $\fracpart{\xbf} = \xbf - \round{\xbf}$ denotes the (centered) fractional part and $\round{\xbf}$ denotes the nearest integer to $\xbf$ with half integers rounded upwards and both $\fracpart{\cdot}$ and $\round{\cdot}$ operate on vectors elementwise.  For $m \geq 2$ the function $\dealias(\xbf)$ can be computed by a simple sequential algorithm \cite[Sec. 7.2.1]{McKilliam2010thesis}.

@@

We now describe the least squares unwrapping (LSU) estimator of the polynomial coefficients. Recall that we desire to estimate the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from the noisy samples $Y_1, \dots, Y_N$ given in~\eqref{eq:Y_nsamplednoisey}.  We take the complex argument of the $Y_n$ and divide by $2\pi$ to obtain
\begin{equation}\label{eq:noise_circ_poly}
\Theta_n = \frac{\angle{Y_n}}{2\pi} = \fracpart{ \Phi_n + y(n) }
\end{equation}
where $\angle{z}$ denotes the complex argument of the complex number $z$, and 
\[
\Phi_n = \frac{1}{2\pi}\angle(1 + \rho^{-1}s_n^{-1}X_n)
\] 
are random variables representing the phase noise induced by the $X_n$~\cite{Tretter1985,Quinn2009_dasp_phase_only_information_loss}.  If the distribution of $X_n$ is circularly symmetric (i.e., the angle $\angle X_n$ is uniformly distributed on $[-\pi, \pi)$ and is independent of the magnitude $\abs{X_n}$) then the distribution of $\Phi_n$ is the same as the distribution of $\tfrac{1}{2\pi}\angle(1 + \rho^{-1}X_n)$.  If the $X_1, \dots, X_N$ are circularly symmetric and identically distributed, then $\Phi_1, \dots, \Phi_n$ are also identically distributed.

@@

It is shown in~\cite{McKilliam2009asilomar_polyest_lattice} and \cite[Sec~8.1]{McKilliam2010thesis} how this minimisation problem can be posed as a nearest lattice point problem, and we briefly discuss this here.  Write $SS$ as
\[
SS(\mubf) = \sum_{n=1}^{N}{ \left( \Theta_{n} - W_n - \sum_{k = 0}^{m}{\mu_k n^k} \right)^2 },
\]
where $W_n = \round{\Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }$ for $n = 1, \dots, N$ are integers called \emph{wrapping variables}.  If we consider $W_1,\dots,W_N$ as nuisance parameters to be estimated then $SS$ can be written as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.  The least squares unwrapping estimator is then found by jointly minimising over $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.   This joint minimisation problem can be solved by computing a nearest point in a lattice.  To see this write $SS$ as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$ using vectors,
\begin{equation}\label{eq:SS_in_vector_form_with_u_nuisance}
SS(\mubf, \wbf) = \| \thetabf - \Xbf\mubf - \wbf \|^2,
\end{equation}
where $\|\cdot\|^2$ denotes the squared Euclidean norm, and $\Xbf$ is the rectangular Vandermonde matrix
\begin{equation}\label{eq:Xvandermonde}
\Xbf = \left[ \begin{array}{ccccc} 
1 & 1 & 1 & \cdots & 1 \\
1 & 2 & 4 & \cdots & 2^m \\
\vdots & \vdots  & \vdots & \ddots & \vdots \\
1 & N & N^2 & \cdots & N^m \\
\end{array} \right],
\end{equation}
and where the column vectors $\wbf = [W_1, \dots, W_N]^\prime$ and $\thetabf = [\Theta_1, \dots, \Theta_N]^\prime$.   For fixed $\wbf$ the minimiser of $SS(\mubf, \wbf)$ with respect to $\mubf$ is
\begin{equation} \label{eq:hat_theta_cal}
\mubf(\wbf) = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime( \thetabf - \wbf) = \Xbf^+( \thetabf - \wbf),
\end{equation}
where $\Xbf^+ = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime$ is the Moore-Penrose pseudoinverse of $\Xbf$.

We may thus compute $\widehat{\mubf}$ as $\widehat{\mubf} = \mu(\widehat{\wbf})$, where $\widehat{\wbf}$ is the minimiser of
\[
SS(\wbf) = SS(\mubf(\wbf),\wbf) = \| \Qbf\thetabf - \Qbf\wbf \|^2,
\]
where $\Qbf = \Ibf - \Xbf\Xbf^+$ and $\Ibf$ is the $N \times N$ identity matrix\footnote{We have slightly abused notation here by reusing $SS$. This should not cause any confusion as $SS(\mubf)$ and $SS(\mubf,\wbf)$ and $SS(\wbf)$ are easily told apart by their inputs.}.

@@

The \emph{intrinsic mean} or \emph{Fr\'{e}chet mean} of $\Phi_n$ is defined as~\cite{McKilliam_mean_dir_est_sq_arc_length2010,bwhk07a,Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005},
\begin{equation}\label{eq:intrmeandefn}
 \mu_{\text{intr}}  = \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2, 
% %&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\phi - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
\end{equation}
and the \emph{intrinsic variance} is
\[
\sigma_{\text{intr}}^2 = E\fracpart{\Theta - \mu_{\text{intr}}}^2 = \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2,
\]
where $\expect$ denotes the expected value.  Depending on the distribution of $\Phi_n$ the argument that minimises~\eqref{eq:intrmeandefn} may not be unique.  The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.  If the minimiser is not unique we say that $\Phi_n$ has no intrinsic mean.  We are now equipped to state the asymptotic properties of the LSU estimator.

@@

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and pdf $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$ then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $(m+1)\times (m+1)$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k = \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}
The proof of Theorem~\ref{thm:asymp_proof} is contained within the next two sections. Section~\ref{sec:strongconstproof} proves strong consistency and Section~\ref{sec:centlimitproof} proves asymptotic normality.  The proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proofs here take a similar approach, but require new techniques.  Before giving the proof we make the following remarks.

>>>>

\begin{proposition}\label{prop:lattice-theory-constructing_a_rectangular_tesselating_region}
Let  $\Lambda$ be an $n$ dimensional lattice and $\Bbf \in \reals^{n\times n}$ be a generator matrix for $\Lambda$. Let $\Bbf = \Qbf\Rbf$ where $\Qbf$ is orthonormal and $\Rbf$ is upper triangular with elements $r_{ij}$.  Then the rectangular prism $\Qbf P$ where
\[
P = \prod_{k=1}^{n}{\left[-\frac{r_{kk}}{2}, \frac{r_{kk}}{2}\right)}
\]
is a set of coset representatives for $\reals^n / \Lambda$.
\end{proposition}
\begin{IEEEproof}
This result is well known~\cite[Chapter IX, Theorem IV]{Cassels_geom_numbers_1997}~\cite[Proposition 2.1]{McKilliam2010thesis}.  %This result is for lattices with full rank.  A result in the general case can be obtained similarly, but is not required here.  
\end{IEEEproof}

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}. The nearest lattice point problem is: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find a lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, solve $\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.$  The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself is considered as an additional input parameter \cite{micciancio_hardness_2001, AjtaiShortestVecProbNPHard1998}.  On the other hand,  fast algorithms that compute a nearest point often exist for specific lattices with known structure~\cite{Conway1982FastQuantDec,McKilliam2009CoxeterLattices,McKilliam2008}.  Even in the general case, where the lattice has no known structure, algorithms exist that can compute a nearest lattice point in reasonable time if the dimension is small \cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. One such algorithm is the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999}.  With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.  Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Yet another approximate algorithm is the $K$-best algorithm studied in~\cite{Zhan2006_K_best_sphere_decoder} that operates similarly to the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} used in coding theory.  The $K$-best algorithm is more expensive that Babai's algorithm, but typically produces a better approximation.  In this paper we make use of the $K$-best algorithm for approximating the LSU estimator.  We will find that it produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for moderate numbers of observations ($N$ less that about 1000).   We will not detail the workings of the $K$-best algorithm in this paper.  A thorough description is given in~\cite{Zhan2006_K_best_sphere_decoder}.

@@

Given a polynomial $g(x) = a_0 + a_1x + \dots + a_m x^m$, let
\[
\coef(g) = \left[ \begin{array}{ccccc} a_0 & a_1 & a_2 & \dots & a_m \end{array} \right]^\prime,
\]
denote the column vector of length $m+1$ containing the coefficients of $g$.  Superscript $^\prime$ indicates the transpose.  If $y$ and $z$ differ by a polynomial from $\mathcal{Z}$ then we can write $y = z + p$ where $p \in \mathcal{Z}$ and then also $\coef(y) = \coef(z) + \coef(p)$.
%\footnote{In group theory terminology $\coef(\cdot)$ coupled with vector addition is called a \term{group homomorphism}.\index{group homomorphism}}. 
  Consider the set
\[
L_{m+1} = \{ \coef(p) \mid p \in \mathcal{Z} \}
\]
containing the coefficient vectors corresponding to the polynomials in $\mathcal{Z}$.  Since the integer valued polynomials are a basis for $\mathcal{Z}$,
\begin{align*}
L_{m+1} &= \{ \coef(c_0 p_0 + c_1p_1 + \dots + c_mp_m) \mid c_i \in \ints \} \\
&= \{ c_0 \coef(p_0) + \dots + c_m\coef(p_m) \mid c_i \in \ints \}.
\end{align*}
Let
\[
\Pbf = \left[ \begin{array}{cccc} \coef(p_0)& \coef(p_1)& \dots& \coef(p_m)  \end{array} \right]
\]
be the $m+1$ by $m+1$ matrix with columns given by the coefficients of the integer valued polynomials.  Then,
\[
L_{m+1} = \{ \xbf = \Pbf\ubf \mid \ubf \in \ints^{m+1} \}
\]
and it is clear that $L_{m+1}$ is an $m+1$ dimensional lattice.  That is, the set of coefficients of the polynomials from $\mathcal{Z}$ forms a lattice with generator matrix $\Pbf$. We can restate Theorem~\ref{thm:circpolysampledthm} as:
\begin{corollary}\label{cor:circpolysampledcoef}
Two polynomial phase signals $s(t) = e^{2\pi j y(t)}$  and $r(t) = e^{2\pi j z(t)}$ are aliases if and only if $\coef(y)$ and $\coef(z)$ differ by a lattice point in $L_{m+1}$.
\end{corollary}

@@

As an example consider the polynomial phase signal of order zero $e^{2\pi j \mu_0}$.  Since $e^{2\pi j \mu_0} = e^{2\pi j(\mu_0 + k)}$ for any integer $k$ we must, in order to ensure identifiability, restrict $\mu_0$ to some interval of length $1$.  A natural choice is the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$. The lattice $L_1$ is the 1-dimensional integer lattice $\ints$ and the interval $[-\nicefrac{1}{2}, \nicefrac{1}{2})$ corresponds to the Voronoi cell of $L_1$. 
When $m=1$ it turns out that a natural choice of identifiable region is the square box $[-\nicefrac{1}{2}, \nicefrac{1}{2})^2$. This corresponds with the \term{Nyquist criterion}.  The lattice $L_2$ is equal to $\ints^2$ and the box $[-\nicefrac{1}{2}, \nicefrac{1}{2})^2$ corresponds with the Voronoi cell of $L_2$.  
When $m > 1$ the identifiable region becomes more complicated and $L_{m+1} \neq \ints^{m+1}$. %However, we can always construct an identifiable region by taking a tessellating region of lattice $L_{m+1}$.  %Two convenient tessellating regions are the Voronoi cell of $L$ and the rectangular tessellating region given by~\eqref{eq:rectangular_identifiable_region}.

In general there are infinitely many choices for the identifiable region. A natural choice is the Voronoi cell of $L_{m+1}$ as used in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}. Another potential choice is a fundamental parallelepiped of $L_{m+1}$. In this paper we will use the rectangular set constructed using Proposition~\ref{prop:lattice-theory-constructing_a_rectangular_tesselating_region}. Observe that $\Pbf$ is upper triangular with $k$th diagonal element equal to $\tfrac{1}{k!}$.  So this rectangular region is
\begin{equation}\label{eq:rectangular_identifiable_region}
B = \prod_{k=0}^{m}\left[ -\frac{0.5}{k!}, \frac{0.5}{k!}  \right).
\end{equation}
We will make use of this region when deriving the statistical properties of the LSU estimator in the next section.

Given vectors $\xbf$ and $\ybf$ in $\reals^{m+1}$ we say that $\xbf \equiv \ybf \bmod L_{m+1}$ if $\xbf$ and $\ybf$ differ by a lattice point in $L_{m+1}$.  We define the function $\dealias(\xbf)$ to take $\xbf$ to its coset representative inside $B$. That is, $\dealias(\xbf) = \zbf \in B$ where $\xbf - \zbf \in L_{m+1}$.  %The effect of $\dealias(\xbf)$ is to \emph{dealias} the polynomial phase coefficients. 
When $m = 0$ or $1$ then $\dealias(\xbf) = \fracpart{\xbf}$ where $\fracpart{\xbf} = \xbf - \round{\xbf}$ denotes the (centered) fractional part and $\round{\xbf}$ denotes the nearest integer to $\xbf$ and both $\fracpart{\cdot}$ and $\round{\cdot}$ operate on vectors elementwise.  The direction of rounding for half-integers is not important so long as it is consistent.  We have chosen to round up half-integers here.  For $m \geq 2$ the function $\dealias(\xbf)$ can be computed by a simple sequential algorithm \cite[Sec. 7.2.1]{McKilliam2010thesis}.

@@

We now describe the least squares unwrapping (LSU) estimator of the polynomial coefficients. Recall that we desire to estimate the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from the noisy samples $Y_1, \dots, Y_N$ given in~\eqref{eq:Y_nsamplednoisey}.  Take the complex argument of the $Y_n$ and divide by $2\pi$ to obtain
\begin{equation}\label{eq:noise_circ_poly}
\Theta_n = \frac{\angle{Y_n}}{2\pi} = \fracpart{ \Phi_n + y(n) }
\end{equation}
where $\angle{z}$ denotes the complex argument of $z$, and 
\[
\Phi_n = \frac{1}{2\pi}\angle(1 + \rho^{-1}s_n^{-1}X_n)
\] 
are random variables representing the phase noise induced by the $X_n$~\cite{Tretter1985,Quinn2009_dasp_phase_only_information_loss}.  If the distribution of $X_n$ is circularly symmetric (i.e., the angle $\angle X_n$ is uniformly distributed on $[-\pi, \pi)$ and is independent of the magnitude $\abs{X_n}$) then the distribution of $\Phi_n$ is the same as the distribution of $\tfrac{1}{2\pi}\angle(1 + \rho^{-1}X_n)$.  If the $X_1, \dots, X_N$ are circularly symmetric and identically distributed, then $\Phi_1, \dots, \Phi_n$ are also identically distributed.

@@

This minimisation problem can be posed as a nearest lattice point problem~\cite{McKilliam2009asilomar_polyest_lattice, McKilliam2010thesis}.  Write $SS$ as
\[
SS(\mubf) = \sum_{n=1}^{N}{ \left( \Theta_{n} - W_n - \sum_{k = 0}^{m}{\mu_k n^k} \right)^2 },
\]
where $W_n = \round{\Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }$ for $n = 1, \dots, N$ are integers called \emph{wrapping variables}.  If we consider $W_1,\dots,W_N$ as nuisance parameters to be estimated then $SS$ can be written as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.  The least squares unwrapping estimator is then found by jointly minimising over $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.   This joint minimisation problem can be solved by computing a nearest point in a lattice.  To see this write $SS$ as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$ using vectors,
\begin{equation}\label{eq:SS_in_vector_form_with_u_nuisance}
SS(\mubf, \wbf) = \| \thetabf - \Xbf\mubf - \wbf \|^2,
\end{equation}
where $\|\cdot\|^2$ denotes the squared Euclidean norm, and $\Xbf$ is the rectangular Vandermonde matrix
\begin{equation}\label{eq:Xvandermonde}
\Xbf = \left[ \begin{array}{ccccc} 
1 & 1 & 1 & \cdots & 1 \\
1 & 2 & 4 & \cdots & 2^m \\
\vdots & \vdots  & \vdots & \ddots & \vdots \\
1 & N & N^2 & \cdots & N^m \\
\end{array} \right],
\end{equation}
and the column vectors $\wbf = [W_1, \dots, W_N]^\prime$ and $\thetabf = [\Theta_1, \dots, \Theta_N]^\prime$.   For fixed $\wbf$ the minimiser of $SS(\mubf, \wbf)$ with respect to $\mubf$ is
\begin{equation} \label{eq:hat_theta_cal}
\mubf(\wbf) = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime( \thetabf - \wbf) = \Xbf^+( \thetabf - \wbf),
\end{equation}
where $\Xbf^+ = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime$ is the Moore-Penrose pseudoinverse of $\Xbf$.  We may thus compute $\widehat{\mubf}$ as $\widehat{\mubf} = \mu(\widehat{\wbf})$, where $\widehat{\wbf}$ is the minimiser of
\[
SS(\wbf) = SS(\mubf(\wbf),\wbf) = \| \Qbf\thetabf - \Qbf\wbf \|^2,
\]
where $\Qbf = \Ibf - \Xbf\Xbf^+$ and $\Ibf$ is the $N \times N$ identity matrix\footnote{We have slightly abused notation here by reusing $SS$. This should not cause any confusion as $SS(\mubf)$ and $SS(\mubf,\wbf)$ and $SS(\wbf)$ are easily told apart by their inputs.}.

@@

The \emph{intrinsic mean} (or \emph{Fr\'{e}chet mean}) of $\Phi_n$ is defined as~\cite{McKilliam_mean_dir_est_sq_arc_length2010,bwhk07a,Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005},
\begin{equation}\label{eq:intrmeandefn}
 \mu_{\text{intr}}  = \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2, 
% %&= \arg \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})}\int_{-\nicefrac{1}{2}}^{\nicefrac{1}{2}}\fracpart{\phi - \mu}^2f(\theta) d\theta. \label{eq:minunrappedmeandef}
\end{equation}
and the \emph{intrinsic variance} is
\[
\sigma_{\text{intr}}^2 = E\fracpart{\Theta - \mu_{\text{intr}}}^2 = \min_{\mu \in [-\nicefrac{1}{2}, \nicefrac{1}{2})} \expect \fracpart{\Phi_n - \mu}^2,
\]
where $\expect$ denotes the expected value.  Depending on the distribution of $\Phi_n$ the argument that minimises~\eqref{eq:intrmeandefn} may not be unique.  The set of minima is often called the~\emph{Fr\'{e}chet mean set}~\cite{Bhattacharya_int_ext_means_2003,Bhattacharya_int_ext_means_2005}.  If the minimiser is not unique we say that $\Phi_n$ has no intrinsic mean.  We are now equipped to state the asymptotic properties of the LSU estimator.

@@

\begin{theorem} \label{thm:asymp_proof} 
Let $\widehat{\mubf}$ be defined by~\eqref{eq:hatmubfLSUdefn} and put $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$.  Denote the elements of $\widehat{\lambdabf}_N$ by $\widehat{\lambda}_{0,N}, \dots, \widehat{\lambda}_{m,N}$.  Suppose $\Phi_1, \dots, \Phi_N$ are independent and identically distributed with zero intrinsic mean, intrinsic variance $\sigma^2$, and pdf $f$, then: 
\begin{enumerate}
\item (Strong consistency) $N^k \widehat{\lambda}_{k,N}$ converges almost surely to $0$ as $N\rightarrow\infty$ for all $k = 0, 1, \dots, m$.
\item (Asymptotic normality) If $f(\fracpart{x})$ is continuous at $x = -\nicefrac{1}{2}$ and $f(-\nicefrac{1}{2}) < 1$ then the distribution of
\[
\left[
\begin{array}
[c]{cccc}%
\sqrt{N} \widehat{\lambda}_{0,N} & N \sqrt{N}\widehat{\lambda}_{1,N}  & \dots & N^m\sqrt{N} \widehat{\lambda}_{m,N}
\end{array}
\right]^\prime
\]
converges to the normal with zero mean and covariance
\[
\frac{\sigma^2}{\left(1-f( -\nicefrac{1}{2}) \right)^{2}} \Cbf^{-1},
\]
where $\Cbf$ is the $(m+1)\times (m+1)$ Hilbert matrix with elements $C_{ik} = 1/(i + k + 1)$ for $i,k \in \{0, 1, \dots, m\}$.
\end{enumerate}
\end{theorem}

A proof of the first part of this theorem (strong consistency) is given in Section~\ref{sec:strongconstproof}.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and for the case when $m=1$ were given in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proof here takes a similar approach, but requires new techniques.  Before giving the proof of strong consistency we make the following remarks.
